Efficient Calculation of Interaction Features on Sparse Matrices

### 2015-08-24 paper/spares_interaction.tex
fixed algorithm and added package files

\usepackage{array}
\usepackage{clrscode3e}
\usepackage{amsmath}

@@

Introduction
Interaction features are a way of capturing correlations between features in a machine 
learning setting. A feature vector $\vec{x}$ of dimensionality $D$ has second degree interaction features 
$\{x_a \cdot x_b : a, b \in \{0,1,..., D-1\} \land a < b\}$, 
so a $D$ dimensional vector has $\binom{D}{2} = \frac{D^2-D}{2}$ second degree interaction features. A naive
approach to calculating these features is to simply iterate through the combinations of the column indices.
For a sparse vector, many of the resulting interaction features would be zero, and could therefore be ignored.
This work describes a method to efficiently calculate second degree interaction features for a sparse matrix 
that has time and space complexities that decrease quadratically with the density of the input matrix with respect to the naive approach.

%Consider the following matrix:

%\[ \left( \begin{array}{cccc}
%3 & 0 & 0 & 3 \\
%0 & 4 & 2 & 0 \\
%1 & 2 & 0 & 3 \end{array} \right)\]

%If each row is an instance vector, then the interaction feature matrix is

%\[ \left( \begin{array}{cccccc}
%0 & 0 & 9 & 0 & 0 & 0 \\
%0 & 0 & 0 & 8 & 0 & 0 \\
%2 & 0 & 3 & 0 & 6 & 0 \end{array} \right)\]

%Note that there are $\binom{D}{2} = \frac{D^2-D}{2}$ columns in the interaction matrix, which is D choose 2, 
%since we generate products of all combinations of 2 features in the original matrix. 
%Which column corresponds with which product pair is not important so long as it’s 
%consistent. In this example, each column in the interaction matrix corresponded to the 
%following product pairs: $(1, 2),  (1, 3),  (1, 4),  (2, 3), ( 2, 4), (3, 4)$.

%Notice that the interaction matrix contains many zero entries. This is of course because 
%the original matrix contained zero entries, so the interaction features, which are 
%products of pairs of features, contain many zeros. This means that the only products that 
%need to actually be calculated or those for which both features in the combination are 
%nonzero. If the original matrix is sparse and represented in a sparse matrix format 
%(e.g. compressed sparse row), a list of nonzero column indices are stored for each row 
%and are easily retrievable in $O(1)$ time. Interaction features can be generated from this 
%list via the following method.

@@

\begin{codebox}
\Procname{$\proc{Sparse Interaction}(A)$}
    \zi $\func{map}(b, a) = \frac{2Da-a^2+2b-3a-2}{2}$
    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2-D}{2}$
    \zi \For $\id{row}$ in $A$ \Do
    \zi     $N_{zc} \gets$ nonzero columns of $row$
    \zi     \For each combination $a, b$ of elements of $N_{zc}$ \Do
    \zi         $k \gets \func{map}(b, a)$
    \zi         $i \gets$ index of $\id{row}$
    \zi         $B[i, k] \gets \id{row}[a] \cdot \id{row}[b]$
            \End
       	\End
\end{codebox}

>>>>

\usepackage{array}
\usepackage{clrscode3e}
\usepackage{amsmath}
\usepackage{kbordermatrix}

@@

Introduction
Interaction features are a way of capturing correlations between features in a machine 
learning setting. A feature vector $\vec{x}$ of dimensionality $D$ has second degree interaction features 
$\{x_i \cdot x_j : i, j \in \{0,1,..., D-1\} \land i < j\}$, 
so a $D$ dimensional vector has $\binom{D}{2} = \frac{D^2-D}{2}$ second degree interaction features. A naive
approach to calculating these features is to simply iterate through the combinations of the column indices.
For a sparse vector, many of the resulting interaction features would be zero, and could therefore be ignored.
This work describes a method to efficiently calculate second degree interaction features for a sparse matrix 
that has time and space complexities that decrease quadratically with the density of the input matrix with respect to the naive approach.

@@

\begin{codebox}
\Procname{$\proc{Sparse Interaction}(A)$}
    \zi $\func{map}(a, b) = \frac{2Da-a^2+2b-3a-2}{2}$
    \zi $N \gets$ row count of $A$
    \zi $D \gets$ column count of $A$
    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2-D}{2}$
    \zi \For $\id{row}$ in $A$ \Do
    \zi     $N_{zc} \gets$ nonzero columns of $row$
    \zi     \For $i \gets 0 \To |N_{zc}|-2$ \Do
    \zi         \For $j \gets i+1 \To |N_{zc}|-1$ \Do
    \zi             $k \gets \func{map}(i, j)$
    \zi             $r \gets$ index of $\id{row}$
    \zi             $B[r, k] \gets \id{row}[i] \cdot \id{row}[j]$
                \End
            \End
       	\End
\end{codebox}
